Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-13 paper.tex
Superscript + (it would be ironic if this IT paper did not even get superscript right)

\title{Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  from the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
  chemists volunteering their time for an Open Science ``movement''---do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion. The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the
drugs we take are the result of intensive and costly research, and the
more information is shared (a process sometimes very hard to achieve
in pharmaceutical companies) the easier it should be to develop
novel treatments. How chemical information relates to computer
science becomes clear when realizing the amount of data available
currently---74 million different molecules are currently stored in
public databases (such as PubChem) with a large amount of annotations,
and the design of data structures for chemicals, as well as subsequent
data mining of this information, are areas where expertise in computer
science is dearly needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

@@

Graph mining algorithms can be applied to this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (although not in others).  Nevertheless, the
3D information is lost when one only considers the graph of a
molecule. To extend the concept of three-dimensionality further, in
many biological systems the \emph{surface properties} of the molecule
are what appear to be relevant, as shown in the bottom right hand
corner.  Hence, while graph mining on chemicals is certainly a
promising avenue to explore, one needs to keep in mind that not all
molecular properties are encoded in the graph (for example, neither
rotational states, nor different charge states are encoded) and thus
often other `descriptors' need to be used to capture the essentials of
a chemical system \cite{bender2004}.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity to collect, store, and mine data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data by bridging the gap between risk in
generalizable encodings and complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

Our motivation for this article is to highlight a variety of topics in
cheminformatics where modern CS research could contribute, and
to present the computational and algorithmic problems
that are addressed in the field.

@@

Most cheminformatics applications rely on large databases of chemical structures
and their properties. Organizing and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
with time several
different SMILES implementations became available with no universal
standard of generating a canonical SMILES. After years of research and various
commercial products, there is today no single public standard for a
line-notation of molecular labeled graphs, which is available to a
broad audience (academic and
commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

@@

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimized for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is completely infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
where c is the count of bits on in the same position in both the two
fingerprints, a is the count of bits on in the first structure, and b is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logic [baaderdl2007]. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas [couto2010]. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

The major difficulty with enumeration is that the \emph{in silico} representation of
molecules, the so-called molecular graphs, are labeled objects (i.e. labeled
graphs), while the atoms of a molecule are of course not uniquely labeled. The
mathematical concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. When enumerating
molecular structures, one has thus to solve the so--called isomorphism problem.
The good news is that isomorphism can be solved in polynomial time for molecular
graphs due to the fact that molecules belong to a restricted class of graphs
known as \emph{bounded valence graphs}; however, that same restriction prevents
applying to a molecular graph the polynomial delay enumeration results obtained
for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that molecules
can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate natural
compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries like GDB-13
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers [3] or tautomers [4]; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
(UCSF) are valuable resources to search for new drugs. For instance
the well known ligand-based design approach uses these
databases along with structure similarity tools to discover new
compounds binding to biological targets (protein or nucleic
acid). Compounds retrieved through ligand-based design, however, must
already be present in databases, which is a limitation of the
technique. To overcome that limitation, algorithms have recently been
developed to enumerate all compounds in the chemical space matching
circular [5] or path fingerprints [6], these fingerprints being
popular tools to measure similarity. Enumerating molecules from
fingerprints, or more generally enumerating molecules from molecular
descriptors, is still very much under development. At stake is the
design of new chemicals or drugs that have not yet been synthesized.

@@

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types, aka compounds. A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

A key aspect of cheminformatics is that much of its development
requires access to collections of molecular structures (and, usually,
associated bioactivity data). As has been noted, it is only recently
that large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale \emph{benchmarking of
algorithms}. The latter is especially relevant for the plethora of
data mining techniques that are employed in cheminformatics. At this
point, the nature of Open Data focuses primarily on structure and
activity data types. On the other hand, there is a distinct lack of
textual data (such as journal articles) that are Open in nature. While
PubMed abstracts serve as a proxy for journal articles, text mining
methods in cheminformatics are hindered by this state of affairs.
Interestingly, patent information is publicly accessible (cf. Google
Patents) and there have been efforts to mine these for chemical
information using Open tools.

Obviously freely accessible data is important for research; but
equally important are the tools to work with chemical data---more
specifically, chemical structures. Chemical toolkits are the workhorses
of cheminformatics research and a variety of toolkits are available,
both commercial (which, in many cases, provide free academic
licenses) and Open Source. In this article we do not debate the
arguments for and against Open Source; rather, we focus on Open Source
tools, since this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

>>>>

\title{Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockvile, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{baoilleach@gmail.com}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  including the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
  chemists volunteering their time for an Open Science ``movement''---do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, bi-protonated, or any variations of those states).
\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (Pubchem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

We should care about chemical information being made public since the
drugs we take are the result of intensive and costly research, and the
more information is shared (a process sometimes very hard to achieve
in pharmaceutical companies) the more easy it should be to develop
novel treatments drugs. How chemical information relates to computer
science becomes clear when realizing the amount of data available
currently---74 million different molecules are currently stored in
public databases (such as PubChem) with a large amount of annotations,
and the design of data structures for chemicals, as well as subsequent
data mining of this information, are areas where expertise in computer
science is dearly needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

@@

Graph mining algorithms can be applied on this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (while not in others).  Nevertheless, the
3D information is lost when one only considers the graph of a
molecule. To extend the concept of three-dimensionality further, in
many biological systems the \emph{surface properties} of the molecule
are what appears to be relevant, as shown in the bottom right hand
corner.  Hence, while graph mining on chemicals is certainly a
promising avenue to explore, one needs to keep in mind that not all
molecular properties are encoded in the graph (for example, neither
rotational states, nor different charge states are encoded) and thus
often other `descriptors' need to be used to capture the essentials of
a chemical system \cite{bender2004}.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity to collect, store, and mine data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data by bridging the gap between risk in
generalizable encodings and complexity (space and time). The goals: to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

Our motivation for this article is to highlight a variety of topics in
cheminformatics and present the computational and algorithmic problems
that are addressed in the field. We hope to
highlight areas where modern CS research could contribute to these
problems.

@@

Most cheminformatics applications rely on large databases of chemical structures
and their properties. Organising and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those challenges poses its unique computer science complexities.

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with one other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES). However, several
different SMILES implementations became available, with no universal
standard of interpretation. After years of research and various
commercial products, there is today no single public standard for a
line-notation of molecular labeled graphs, which is available to a
broad audience (academic and
commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

@@

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimised for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is completely infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbours, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
where c is the count of bits on in the same position in both the two
fingerprints, a is the count of bits on in the first structure, and b is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logic [baaderdl2007]. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas [couto2010]. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

The major difficulty with enumeration is that the \emph{in silico} representation of
molecules, the so-called molecular graphs, are labeled object (i.e. labeled
graphs), while the atoms of a molecule are of course not uniquely labeled. The
mathematical concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. When enumerating
molecular structures, one has thus to solve the so--called isomorphism problem.
The good news is that isomorphism can be solved in polynomial time for molecular
graphs due to the fact that molecules belong to a restricted class of graphs
known as \emph{bounded valence graphs}; however, that same restriction prevents
applying to a molecular graph the polynomial delay enumeration results obtained
for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that molecules
can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate natural
compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries like GDC-13
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers [3] or tautomers [4]; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
(UCSF) are valuable resources to search for new drugs. For instance
the well known ligand-based design approach is making use of these
databases along with structure similarity tools to discover new
compounds binding to biological targets (protein or nucleic
acid). Compounds retrieved through ligand-based design, however, must
already be present in databases, which is a limitation of the
technique. To palliate that limitation, algorithms have recently been
developed to enumerate all compounds in the chemical space matching
circular [5] or path fingerprints [6], these fingerprints being
popular tools to measure similarity. Enumerating molecules from
fingerprints, or more generally enumerating molecules from molecular
descriptors, is still very much under development. At stake is the
design of new chemicals or drugs that have not yet been synthesized.

@@

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types, aka compounds. A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like Chembl and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

A key aspect of cheminformatics is that much of its development
requries access to collections of molecular structures (and, usually,
associated bioactivity data). As has been noted, it is only recently
that large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale \emph{benchmarking of
algorithms}. The latter is especially relevant for the plethora of
data mining techniques that are emoloyed in cheminformatics. At this
point, the nature of Open Data focuses primarily on structure and
actvity data types. On the other hand, there is a distinct lack of
textual data (such as journal articles) that are Open in nature. While
PubMed abstracts serve as a proxy for journal articles, text mining
methods in cheminformatics are hindered by this state of affairs.
Interestingly, patent information is publicly accessible (cf. Google
Patents) and there have been efforts to mine these for chemical
information using Open tools.

Obviously freely accessible data is important for research; but
equally important are the tools to work with chemical data---more
specifically, chemical structures. Chemical toolkits are the workhorses
of cheminformatics research and a variety of toolkits are available,
both commericial (which, in many cases, provide free academic
licenses) and Open Source. In this article we do not debate the
arguments for and against Open Source; rather, we focus on Open Source
tools, since this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.
